---
title: 'Notes 03: Multiple Linear Regression'
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, 
                      comment = "", 
                      message = FALSE, 
                      warning = FALSE)

#mosaic package makes some functions more user-friendly for Intro Stat topics
library(mosaic)

#ggformula package is for graphics
library(ggformula)

#janitor package will help us clean up variable names
library(janitor)

#for other data management 
library(tidyverse)

#for scatterplot matrix
library(lattice) 

#for VIF calculation
library(car) 

#for parallel slopes graph
library(moderndive)

#customize graph theme for full document
theme_set(theme_light())
```


## Cigarette Example
Recall from our correlation analysis for the cigarette data, that we observed the correlations between all variables:
```{r}
#| label: data
cigs <- read_csv("cigarettes.csv")

#clean up variable names
cigs <- cigs |>
  clean_names()
```

```{r}
#| label: corr
#look a scatterplots and correlations between variables
cor(cigs[,2:5])
pairs(cigs[,2:5])
```

We can also use the library `lattice` to create the scatterplots.
```{r}
#| label: lattice
#Putting "lattice::" before the function name just directly references which package the function is a part of
lattice::splom(cigs[,2:5])
```


Suppose we were to construct a model with all variables for CO output.
```{r}
#| label: model
#construct the model
cig_mod <- lm(co ~ tar + nicotine + weight, data = cigs)

#output coefficients
coef(cig_mod)
```


We then evaluate the model:
```{r}
#| label: tests
#remember both the F and t-test results are included in this output
summary(cig_mod)
```
Despite the strong relationship between Nicotine and CO, we do not find the variable to be a statistically significant predictor given the other variables in the model.

Why is this?




Let's observe the the Variance Inflation Factor associated with our model.  You will need the package `car` to calculate the VIF

```{r}
#| label: vif
#vif function comes from the "car" package that we loaded at the top
vif(cig_mod)
vif(cig_mod) > 10
```

```{r}
#| label: assumptions
#create diagnostic plots for assumptions
plot(cig_mod, 1, add.smooth = FALSE)
plot(cig_mod, 2)
plot(cig_mod, 5, add.smooth = FALSE)
```


Notice that Tar and Nicotine both have a very high VIF associated with them, indicating that they are most likely collinear.  Here is how the VIF is calculated

```{r}
#| label: vif_calc
#CO is not involved, find the R^2 associated with a model for Nicotine
nic_mod <- lm(nicotine ~ weight + tar, data = cigs) 

#extracts the NOT adjusted R^2
(nicr_2 <- summary(nic_mod)$r.squared)

#calculates the VIF for nicotine
nic_vif <- 1/(1 - nicr_2) 
nic_vif

```
So we cannot include both Nicotine and Tar in our model - how do we choose which to include?  Based on the model output, which model would you choose?
```{r}
#| label: model_nic
#linear model including nicotine
cig_mod_nic <- lm(co ~ nicotine + weight, data = cigs)
summary(cig_mod_nic)

#diagnostic plots
plot(cig_mod_nic, 1, add.smooth = FALSE)
plot(cig_mod_nic, 2)
plot(cig_mod_nic, 5, add.smooth = FALSE)
```

```{r}
#| label: model_tar
#linear model including tar
cig_mod_tar <- lm(co ~ tar + weight, data = cigs)
summary(cig_mod_tar)

#diagnostic plots
plot(cig_mod_tar, 1, add.smooth = FALSE)
plot(cig_mod_tar, 2)
plot(cig_mod_tar, 5, add.smooth = FALSE)
```

## Height and Weight by Sex

Let's observed how we might include a categorical variable in our model.
```{r}
#| label: davis_data
data(Davis) #found in the car package

#clean names
davis <- Davis |>
  clean_names()

#Just view the first couple of rows of data
head(davis) 

#variable information
str(davis)
```


We can observe the relationship between height and weight for the students using a plot, but does that provide all the information we need?  Also, what do you notice from the plot?

```{r}
#| label: davis_scat
#not including sex in the model
gf_point(weight ~ height, data = davis, 
         xlab = "Height (cm)", ylab = "Weight (kg)")
```


There is an obvious outlier and after some investigation, we see it is row 12.  Therefore, we want to remove that data point from our analysis (this is most likely a data entry error as there would not be someone who is less than 60 cm in real life, or at least not an adult).
```{r}
#| label: no_outlier
new_davis <- davis[-12,] #removes row 12 from the dataframe

new_davis <- davis |> #tidyverse method for removing outliers
  filter(height > 60)
```


Let's use some new arguments in `ggformula` to look at our data, but split by sex.
```{r}
#including sex in the model, WITH outlier
gf_point(weight ~ height, data = davis, 
         shape = ~sex, color = ~sex, 
         xlab = "Height (cm)", ylab = "Weight (kg)") |>
  gf_lm(size = 1) +
  ylim(40,180) +
  scale_color_viridis_d(end = 0.85)

#including sex in the model, outlier removed
gf_point(weight ~ height, data = new_davis, 
         shape = ~sex, color = ~sex, 
         xlab = "Height (cm)", ylab = "Weight (kg)") |>
  #linear model graph uses parameters from gf_point: 2 separate lines
  gf_lm(size = 1) +
  scale_color_viridis_d(end = 0.85)
```


By default, the above code treats the lines for each sex independently. We can force the lines to have identical slopes with the code below:

```{r}
gf_point(weight ~ height, data = new_davis, 
         shape = ~sex, color = ~sex, 
         xlab = "Height (cm)", ylab = "Weight (kg)") +
  #geom_parallel_slopes comes from moderndive package
  geom_parallel_slopes(se = FALSE) +
  scale_color_viridis_d(end = 0.85)
```


Here are the model coefficients

```{r}
#| label: factor
#factor tells R to treat the "sex" variable as categorical
#Run the model, view coefficients and hypothesis tests
model_hws <- lm(weight ~ height + factor(sex), data = new_davis)
coef(model_hws)
```

Hypohtesis Testing

```{r}
#| label: davis_tests
#hypothesis tests, coefficients, and VIF calculations
summary(model_hws)
vif(model_hws)
```


What are the additional assumptions we are making to construct a model with a qualitative variable?


```{r}
#| label: hws_assumptions
#diagnostic plots for the hws model
plot(model_hws, 1, add.smooth = FALSE)
plot(model_hws, 2)
plot(model_hws, 5, add.smooth = FALSE)
```



```{r}
#| label: mod_interaction
#factor tells R to treat the "sex" variable as categorical
#Run the model, view coefficients and hypothesis tests
model_w_hs_int <- lm(weight ~ height*factor(sex), data = new_davis)
coef(model_w_hs_int)
```

```{r}
#| label: int_tests
#hypothesis tests and coefficients for interaction model
summary(model_w_hs_int)

#vif calculations
vif(model_w_hs_int)
```
What are the additional assumptions we are making to construct a model with a qualitative variable?


```{r}
#| label: int_assumptions
#diagnostic plots for interaction model
plot(model_w_hs_int, 1, add.smooth = FALSE)
plot(model_w_hs_int, 2)
plot(model_w_hs_int, 5, add.smooth = FALSE)
```

