---
title: "Multiple Regression Examples - MEANS"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", message = FALSE, warning = FALSE)

#mosaic package makes some functions more user-friendly for Intro Stat topics
library(mosaic)

#ggformula package is for graphics
library(ggformula)

#janitor package will help us clean up variable names
library(janitor)

#car package for VIF calculations
library(car)

#general data cleaning and management
library(tidyverse)

#dataset
library(palmerpenguins)

#for parallel slopes model
library(moderndive)


theme_set(theme_light())
```


## Read Data

First, let's create the summary statistics and graphics for our comparison.
```{r}
#from the palmerpenguins package
data(penguins)
str(penguins)

#names are already clean, don't technically need this
penguins <- penguins |>
  clean_names()
```

Note that `species` is recorded as a Factor variable (categorical variable with a set number of levels). This means that R will recognize `species` as categorical automatically.

## Exploratory Data Analysis


```{r}

#boxplot of bill length for each species
gf_boxplot(bill_length_mm ~ species, data = penguins,
           ylab = "Bill Length (mm)",
           xlab = "Penguin Species")
```



```{r}
#R already know that Species is categorical because of how it is saved. If you need to specify that a variable be treated as categorical, you can use "factor" as seen below:
gf_boxplot(bill_length_mm ~ factor(species), data = penguins,
           ylab = "Bill Length (mm)",
           xlab = "Penguin Species")
```


```{r}
#summary stats
df_stats(bill_length_mm ~ species, data = penguins)
mean(bill_length_mm ~ species, data = penguins, na.rm = TRUE)
sd(bill_length_mm ~ species, data = penguins, na.rm = TRUE)

```

## Model


First, let's just compare the Bill Length among the three species.  Exactly what comparisons are being made with this model?  Have we done all pairwise comparisons?
```{r}
#create model
model_bill <- lm(bill_length_mm ~ factor(species), data = penguins)

#output coefficients
coef(model_bill)
```

## Inference


```{r}
#Hypothesis Test
summary(model_bill)

```

## Change Reference Level

How do we change the reference level if we want to, so that the default is not alphabetical order? Here are two options. Choose which one you like better!

Option 1:
```{r}

#remember you only NEED to include the factor function if your variable is not already classified as categorical in R. 
#If you want to reorder the levels to change the reference, this can be done within the factor function
#within the factor function, you must list the order that you want all the levels to have. This gives you the most control over the order 

penguins <- penguins |> 
  mutate(species = factor(species, 
                          levels = c("Gentoo", "Adelie", "Chinstrap")))

model_bill_rl <- lm(bill_length_mm ~ factor(species), 
                    data = penguins)
coef(model_bill_rl)
```

Option 2:
```{r}


#alternatively, you can use the "relevel" function to just specify the reference level
#you might need to use the factor function in addition to relevel if the variable is recorded as numerical
#Change the reference level

penguins$species <- relevel(penguins$species, ref = "Gentoo")


model_bill_rl <- lm(bill_length_mm ~ relevel(species, ref = "Gentoo"), data = penguins)
coef(model_bill_rl)
```

## CI


Create confidence intervals for the predicted population mean
```{r}
predict(model_bill, interval = "confidence", newdata = data.frame(species = "Chinstrap"))
```

## Assumptions

```{r}
plot(model_bill_rl, 1)
plot(model_bill_rl, 2)
plot(model_bill_rl, 5)
```


## Larger MLR

Let's observe larger models for Bill Length.  


```{r}
model_2 <- lm(bill_length_mm ~ factor(species) + bill_depth_mm, data = penguins)
coef(model_2)

vif(model_2)
```


Create the graph with parallel slopes from the ModernDive package function
```{r}

gf_point(bill_length_mm ~ bill_depth_mm, data = penguins,
         color = ~factor(species)) +
  geom_parallel_slopes(se = FALSE) +
  scale_color_viridis_d(end = 0.85) +
  labs(x = "Bill Depth (mm)",
       y = "Bill Length (mm)",
       color = "Species of Penguin")
```

Or, graph the regression lines with parallel lines by extracting the coefficients from the model.

```{r}
#to create a plot with three parallel lines, we have to do a bit calculations to get the predicted values with the same slope and then plot those lines.
penguins_predict <- penguins |> 
  drop_na(species, bill_length_mm, bill_depth_mm) |> 
  mutate(predicted_bill_length = predict(model_2))


gf_point(bill_length_mm ~ bill_depth_mm, data = penguins_predict,
         color = ~factor(species)) |> 
  gf_line(predicted_bill_length ~ bill_depth_mm) +
  scale_color_viridis_d(end = 0.85) +
  labs(x = "Bill Depth (mm)",
       y = "Bill Length (mm)",
       color = "Species of Penguin")
```

Another model

```{r}
model_3 <- lm(bill_length_mm ~ factor(species) + bill_depth_mm + flipper_length_mm + body_mass_g,
              data = penguins)
coef(model_3)
vif(model_3)
```

