---
title: "Notes 08: Multiple Poisson Regression and Rates"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
#| echo: false
knitr::opts_chunk$set(echo = TRUE, comment = "")

# standard packages
library(mosaic)
library(ggformula)
library(janitor)
```


## Multiple Poisson Regression - Horseshoe Crabs

### Read in Data

```{r}
crab <- read.table("crab.txt")
```

The data does not come with column names so we will make the column.
```{r}
colnames(crab) <- c("obs", "color", "spine", "width", "weight", "sat")

str(crab)
head(crab)
```


### Multiple Poisson Model

It is possible that the overdispersion issue is related to a need to include more variables in our model.  Let's consider the inclusion of "color" into the model.  Color is a categorical variable with four levels.  Therefore, we need to set the baseline.  We will use color 4 (dark) as our baseline:

```{r}
# create a new dataset where we change reference levels
crab_ref <- crab |>
  # change color to categorical, and make reference level dark (the fourth level)
  mutate(color = factor(color, levels = c("4", "1", "2", "3"))) |>
  # rename the category levels to something more memorable
  mutate(color = recode_factor(color, "4" = "dark", "1" = "lightmed", "2" = "med", "3" = "darkmed"))
```

Now we can build our model with Color and Width:
```{r}
model_cw <- glm(sat ~ width + color, family = poisson(link = log), data = crab_ref)
summary(model_cw)
```

Let's check for overdispersion:
```{r}
model_cw$deviance / model_cw$df.residual
1 - pchisq(model_cw$deviance, model_cw$df.residual)
```

Our model appears to be overdispersed, so let's fit our model using the quasipoisson model.
```{r}
model_cw_disp <- glm(sat ~ width + color, family = quasipoisson(link = log), data = crab_ref)
summary(model_cw_disp)
summary(model_cw_disp)$dispersion
```

Though color is a single variable, we treat it as three dummy variables.  Therefore, to test it specifically, we want to perform a Likelihood Ratio Test/Drop-in-Deviance Test.  
$H_0: \beta_{c1}=\beta_{c2}=\beta_{c3}=0$  
$H_1:$ at least one $\beta_{ci} \neq 0$

As a reminder, we need the output from the model with just width and no color as a predictor. 
```{r}
model_disp <- glm(sat ~ width, data = crab_ref, family = quasipoisson(link = log))
summary(model_disp)
```

Conduct the drop in deviance test.
```{r}
anova(model_disp, model_cw_disp, test = "Chisq")
```
What did we find out?



Finally, we can add in the other variables to the model:
```{r}
mod_full <- glm(sat ~ width + weight, family = quasipoisson(link = log), data = crab_ref)
summary(mod_full)
```

Why are none of the variables significant?  We still need to check for multicollinearity!

```{r}
# scatterplot of weight vs width
gf_point(width ~ weight,
  data = crab_ref,
  xlab = "Weight",
  ylab = "Width"
)

# calculate VIF
library(car)
vif(mod_full)
```



## Modeling Rates instead of Counts

### Credit Card Data

Here we have data on credit card occurrences by income.  In order to fit the model, we need to create a column of the log(cases).
```{r}
credit_card <- read.csv("creditcard.csv")
colnames(credit_card)

credit_card <- credit_card |>
  clean_names() |>
  # creates new column for the log cases per count and income
  mutate(l_cases = log(cases))

head(credit_card)
```
Interpret the third line of data in the output.


### Model for Rates

Now let's fit our model.  Note that the summary output does NOT show anything about the offset, we just have our usual beta parameter estimates.
```{r}
log_fit <- glm(cr_cards ~ income + offset(l_cases), family = poisson, data = credit_card)
summary(log_fit)
```

Here is an alternative way to write the offset statement:
```{r}
log_fit2 <- glm(cr_cards ~ income, offset = l_cases, family = poisson, data = credit_card)
summary(log_fit2)
```

Interpret the slope in context:
```{r}
exp(coef(log_fit)[2])
```

What if we had fit the model without the offset?
```{r}
log_fit_no <- glm(cr_cards ~ income, family = poisson, data = credit_card)
summary(log_fit_no)
```


What if we had not modeled based on rate but count?  First we would have to reorganize the data so that each row represents an individual instead of a group.  Is this the best way to model the data though? (Remember our data is grouped by income).
```{r}
# may need to install, allows us to use uncount()
library(tidyr)
# create new dataset with one row per individual
ind_credit <- uncount(credit_card, weight = cases)

# run model for counts
count_mod <- glm(cr_cards ~ income, family = poisson, data = ind_credit)
summary(count_mod)
```
