---
title: "Notes 06: Validation and Cross-Validiation"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false

---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, comment = "", message = FALSE, warning = FALSE)

#install this package first (R package for Introduction to Statistical Learning Book)
library(ISLR)

library(ggformula)
library(mosaic)

```


## Validation
We will be using data from the `ISLR` package.
```{r}
#| label: auto_data
#load the Auto dataset from the ISLR package
data(Auto)
```

We can observe the structure of the data using the `str()` and `head()` functions.
```{r}
#| label: data_str
str(Auto)
head(Auto)
```

1. How many variables are in the Auto dataset?  

>Insert Answer Here  


2. How many observations are in this dataset?  
```{r}
#| label: row_count
#total observations
n <- nrow(Auto)
```

>Insert Answer Here  

3. We want to predict a car’s miles-per-gallon (`mpg`) using `horsepower`. Create the scatterplot of these two variables and be sure to appropriately label the graph. What sort of relationship do you see?    

```{r}
#| label: scatter_hp
gf_point(mpg ~ horsepower,
         data = Auto,
         xlab = "Horsepower",
         ylab = "Miles Per Gallon")
```

>Insert Answer Here  


### Selecting a Training Set
We need to choose a subset of the data to be our training set.  Let's start with a 50% training set.  The easiest way to do this is to choose a subset of 50% of the *row id numbers* so that we can easily extract all columns of data associated with each row.  Rerun the code chunk - does the training data remain the same each time?  How does the training data compare to the original data?    

```{r}
#| label: select_training
#sample (without replacement) half of the rows 

#this code says: out of n numbers, pick half of them
train <- sample(n, (0.50) * n, replace = FALSE)
#view all the row numbers you have selected but sorted in order
sort(train)
```


```{r}
#| label: scatter_training
#plot JUST training data
Auto[train, ] %>%
  gf_point(mpg ~ horsepower,
           xlab = "Horsepower",
           ylab = "Miles Per Gallon")

```

>Insert Answer Here


Now rerun the code, but first run the code `set.seed(1)`.  Rerun the code chunk again.  Do the selected rows change?
```{r}
#| label: seed_training
set.seed(1)
#sample (without replacement) half of the rows 
#this code says: out of n numbers, pick half of them
train <- sample(n, (0.50) * n, replace = FALSE)
#view all the row numbers you have selected but sorted in order
sort(train)
#plot the training data
Auto[train, ] %>%
  gf_point(mpg ~ horsepower,
           xlab = "Horsepower",
           ylab = "Miles Per Gallon")
```


4.  What does the `set.seed()` function do?  

>Insert Answer Here  

5. How many rows of data will there be in our training set? What about in our testing set?   

>Insert Answer Here  



### Fitting the Model to the Training Set
	

6.  Use the code to fit the first-order linear model using the training data. How does this line of code look different from previous times that we have used the `lm()` function?  

```{r}
#| label: training_model_1
#subset option in lm will only use the rows of data you specify to fit the model
#fit the model ONLY using the training data
lm_fit <- lm(mpg ~ horsepower, data = Auto, subset = train)
summary(lm_fit)
```

7. Is the model significant? What is the value of Adjusted $R^2$?  

>Insert Answer Here  


8. Run the following line of code.  
```{r}
#| label: mse_testing
mean((Auto$mpg - predict(lm_fit, Auto))[-train] ^ 2)
```


8a. Explain in words what each piece of this code is doing. (If it helps, try running small pieces of this line of code at a time, such as just `predict(lm_fit, Auto)` to start understanding what all the pieces do).  

>Insert Answer Here  



8b. What is the value you get from this code?  

>Insert Answer Here  



### Fitting Other Models to the Training Data  

9. We noticed a non-linear relationship between `mpg` and `horsepower`.  We want to make sure we fit the appropriate model before assess its fit via validation methods. Now fit the quadratic model, saved as `lm_fit2`.  

```{r}
#| label: training_model_2
lm_fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train)
summary(lm_fit2)
mean((Auto$mpg - predict(lm_fit2, Auto))[-train] ^ 2)
```

9a. What does the `poly()` function do?  

>Insert Answer Here  


9b. What is the value of the test MSE?  

>Insert Answer Here  


10. Now fit the cubic model, `lm_fit3`. What is the value of the test MSE?
```{r}
#| label: training_model_3
lm_fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, subset = train)
summary(lm_fit3)
mean((Auto$mpg - predict(lm_fit3, Auto))[-train] ^ 2)
```

> Insert Answer Here

11. Of these three models (first-order, quadratic, and cubic), which has the smallest test MSE? What does this mean in terms of how well each model predicts?  

>Insert Answer Here  


12. Remember that we **randomly** split the data into the testing and training groups. Now, use `set.seed(2)` to split the data. Are the same observations put into each group as when we used `set.seed(1)`?

```{r}
#| label: new_training
set.seed(2)
train <- sample(n, (0.50) * n, replace = FALSE)
sort(train)
```

>Insert Answer Here  


13. After using `set.seed(2)`, rerun each of the 3 linear models and recalculate the test MSE’s. What do you notice compared to the previous set of 3 test MSE’s?

```{r}
#| label: new_training_models
lm_fit <- lm(mpg ~ horsepower, data = Auto, subset = train)
mean((Auto$mpg - predict(lm_fit, Auto))[-train] ^ 2)

lm_fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train)
mean((Auto$mpg - predict(lm_fit2, Auto))[-train] ^ 2)

lm_fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto, subset = train)
mean((Auto$mpg - predict(lm_fit3, Auto))[-train] ^ 2)
```

>Insert Answer Here  


## Cross-Validation
First, in order to do cross-validation we cannot use the `lm()` function to fit our regression models.  Instead, we will use the function `glm()`.  The difference between these two functions is that `lm()` uses least squares regression and `glm()` uses a method call maximum likelihood, which we will discuss later.  In most cases, if conditions are satisfied for `lm()` the `glm()` method will produce the same model fit.

1. Run the model using both functions and compare the output. What is the same? What is different?

```{r}
#| label: glm_vs_lm
#reminder of  lm model output (using all data)
lm_fit <- lm(mpg ~ horsepower, data = Auto)
summary(lm_fit)

#glm model
glm_fit1 <- glm(mpg ~ horsepower, data = Auto)
summary(glm_fit1)
```

>Insert Answer Here  



2. Consider the following code to run a K-Fold Cross-Validation. Why do we need to set the seed in this code?  

```{r}
#| label: cv_linear
#Use the cv.glm function from the "boot" library
#(Make sure to install the package!)
library(boot)
set.seed(17)

#Calculate cross-validation prediction error
cv.err1 <- cv.glm(Auto, glm_fit1, K = 5)

#This code outputs the average Test MSE
cv.err1$delta[1]
```

>Insert Answer Here  


3. Run the cv.glm code. How many “folds” or groups is this code creating? Did this code require the model from the lm function or from the glm function?  

>Insert Answer Here  


4. What value did you get for the average Test MSE?  

>Insert Answer Here  

Now use the quadratic model and for the K-Fold Cross-Validation.  
```{r}
#| label: cv_quad
glm_fit2 <- glm(mpg ~ poly(horsepower, 2), data = Auto)
cv.err2 <- cv.glm(Auto, glm_fit2, K = 5)
cv.err2$delta[1]
```


5.  What is your average Test MSE?  

>Insert Answer Here  

Now use the cubic model and for the K-Fold Cross-Validation.
```{r}
#| label: cv_poly3
glm_fit3 <- glm(mpg ~ poly(horsepower, 3), data = Auto)
cv.err3 <- cv.glm(Auto, glm_fit3, K = 5)
cv.err3$delta[1]
```

6. What is your average Test MSE?  

>Insert Answer Here  


7. Which of the three models would you recommend as “best” based on the results?  

>Insert Answer Here  


8. Now look at the code that uses a “for” function.  What do you think this code is doing?  
```{r}
set.seed(20)
cv.err <- rep(0, 5)
#system.time will keep track of how long the computation takes
system.time(for (i in 1:5) {
  glm_fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.err[i] <- cv.glm(Auto, glm_fit, K = 5)$delta[1]
})

plot(cv.err, type = "b")
```

>Insert Answer Here  


### Leave-One-Out Cross-Validation
For LOOCV, either set `k=` to sample size, or remove the `k=` argument entirely.  Adapt the provided code associated with the `for` loop.  

```{r}
#| label: loop_k1_cv
set.seed(13)
cv.err <- rep(0, 5)
#system.time will keep track of how long the computation takes
system.time(for (i in 1:5) {
  glm_fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
  cv.err[i] <- cv.glm(Auto, glm_fit)$delta[1]
})


plot(cv.err, type = "b")
cv.err
```

9.	Compare how long the loop takes to run for K=5 versus using LOOCV (you will have to modify the code for LOOCV). What is the difference?  

>Insert Answer Here  


10.	Compare the values in the cv.err output. Which size model has the LOWEST Test MSE? Do you think this model is actually the best one? Why or why not?  

>Insert Answer Here  


