---
title: "Simple Logistic Regression"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", warning = FALSE, message = FALSE)
# standard packages we use
library(mosaic)
library(ggformula)
library(janitor)

# for data cleaning
library(dplyr)
```

## Simple Logisitic Regression

There are two ways data may be organized for Logistic Regression.  Here is an example where the response data is organized by counts.  Data may take this form when the explanatory variables have distinct grouping values (categorical or discrete) and NOT when explanatory variables are continuous.  
```{r}
# Logistic Regression for ck data
count_ha <- read.csv("countHeartAttack.csv")

count_ha <- count_ha |>
  clean_names()

str(count_ha)


count_ha <- count_ha |> 
  mutate(total = ha + no_ha)

```

To fit our model, we need to define the two columns that represent Y=1 and Y=0 as our response using `cbind()`.  
```{r}
log_ha <- glm(cbind(ha, no_ha) ~ ck, 
              family = binomial(link = "logit"), data = count_ha)
summary(log_ha)
```


What happens if we change the order of the variables in the `cbind()` function?
```{r}
log_no_ha <- glm(cbind(no_ha, ha) ~ ck, family = binomial(link = "logit"), data = count_ha)
summary(log_no_ha)
```


### Likelihood Ratio Test
Note that the deviance values are different than in lecture notes, but the G statistic is the same.  The output in the notes used the individual data, not the count data. You can "hand calculate" it or use `anova()` to calculate the LRT. 
```{r}
# Calculate G-statistic
G <- log_ha$null.deviance - log_ha$deviance
# 311.29-28.14

# degrees of freedom for test
df <- log_ha$df.null - log_ha$df.residual

# p-value
1 - pchisq(G, df)
anova(log_ha, test = "Chisq")
```


### Logistic Regression with Individual Data
Here is an example of the logistic model and LRT for when data is formatted in a "tidy" format - each row a person and each column a variable. Notice: for Individual vs Count data setup, the G statistics, degrees of freedom, and p-values are the same. BUT, the actual Null and Residual Deviances are Different!
```{r}
# read in the data
ind_ha <- read.csv("indHeartAttack.csv")

ind_ha <- ind_ha |>
  clean_names()

str(ind_ha)
```

```{r}
# run the model
log_ha_ind <- glm(ha ~ ck, data = ind_ha, family = binomial)
summary(log_ha_ind)
```


```{r}
# G stat for ck model with Individual Data
# 470.92-187.77
G <- log_ha_ind$null.deviance - log_ha_ind$deviance

# degrees of freedom
# 2 parameters in full model, 1 in reduced
df <- log_ha_ind$df.null - log_ha_ind$df.residual

# p-value
1 - pchisq(G, df)

# automated result
anova(log_ha_ind, test = "Chisq")
```

## Student sleep study example
Here is the Logistic Model
```{r}
sleep <- read.csv("sleep.csv")
model <- glm(hours ~ ages, family = binomial("logit"), data = sleep)
summary(model)
```

Here is the LRT
```{r}
# G statistic
G <- model$null.deviance - model$deviance
# 569.60-566.19

# degrees of freedom
df <- model$df.null - model$df.residual

# p-value
1 - pchisq(G, df)
# Fail to reject the null hypothesis
# Another way to get the same test
anova(model, test = "Chisq")
```

Since there are only five values for the explanatory variable, we can summarize the proportion of "successes" per age group to visually observe our model fit.  Again, if our explanatory variable is continuous, this would not be as straightforward.  

```{r}
# group_by and summarise from the dplyr package, loaded at beginning
prophours <- sleep |>
  group_by(ages) |>
  summarise(prop = mean(hours))

# scatterplot
gf_point(prop ~ ages,
  data = prophours,
  xlab = "Age (years)",
  ylab = "Proportion of Students") |>
  gf_lm()

# predictions
pred <- predict(model, newdata = data.frame(ages = 14:18), type = c("response"))
cbind(prophours, pred)
```
