---
title: "Multiple Logistic Regression"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false

---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", warning = FALSE, message = FALSE)

# standard packages we use
library(mosaic)
library(ggformula)
library(janitor)
```

## Multiple Logisitic Regression
We can extend everything we know about Logistic Regression and Multiple Regression in general to include multiple explanatory variables to predict a binary outcome.

```{r}
library(HSAUR) # be sure to install, this package has the data plasma
data(plasma)

plasma <- plasma |>
  clean_names()

full <- glm(esr ~ fibrinogen + globulin, family = binomial(link = "logit"), data = plasma)

# odds ratios for partial slopes and intercept
exp(coef(full))

# probability when x=0
exp(coef(full)[1]) / (1 + exp(coef(full)[1]))

# significance tests
summary(full)
```

Just like with multiple regression, we need to test the full model, or the general hypothesis.  We can do this with the G-test.

$H_0: \beta_1 = \beta_2=0$  
$H_1:$ at least one $\beta_i \neq 0$  

```{r}
# G statistic
# 30.885-22.971
G <- full$null.deviance - full$deviance

# degrees of freedom
df <- full$df.null - full$df.residual

# p-value
1 - pchisq(G, df)
```


## Nested Model Tests
Just like with multiple regression and the Nested F-test, we can test for subsets of variables or individual variables using the G-test.
```{r}
# fit model and print significance tests
reduced <- glm(esr ~ fibrinogen, family = binomial(link = "logit"), data = plasma)
summary(reduced)

# Testing if Gam can be removed from the model
# 24.84-22.971
G <- reduced$deviance - full$deviance
df <- reduced$df.residual - full$df.residual
1 - pchisq(G, df)

# automated way to get p-value from the nested test
anova(reduced, full, test = "Chisq")
```
