---
title: "ROC Curves"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", warning = FALSE, message = FALSE)
```

## Multiple Logisitic Regression
We can extend everything we know about Logistic Regression and Multiple Regression in general to include multiple explanatory variables to predict a binary outcome.

```{r}
library(HSAUR) # be sure to install, this package has the data plasma
data(plasma)

full <- glm(ESR ~ fibrinogen + globulin, family = binomial(link = "logit"), data = plasma)
exp(coef(full)) # odds ratios for partial slopes and intercept
exp(coef(full)[1]) / (1 + exp(coef(full)[1])) # probability when x=0
summary(full)
```

Just like with multiple regression, we need to test the full model, or the general hypothesis.  We can do this with the G-test.

$H_0: \beta_1 = \beta_2=0$  
$H_1:$ at least one $\beta_i \neq 0$  

```{r}
# 30.885-22.971
(G <- full$null.deviance - full$deviance)
(df <- full$df.null - full$df.residual)
1 - pchisq(G, df)
```


## Nested Model Tests
Just like with multiple regression and the Nested F-test, we can test for subsets of variables or individual variables using the G-test.
```{r}
reduced <- glm(ESR ~ fibrinogen, family = binomial(link = "logit"), data = plasma)
summary(reduced)

# Testing if Gam can be removed from the model
# 24.84-22.971
(G <- reduced$deviance - full$deviance)
(df <- reduced$df.residual - full$df.residual)
1 - pchisq(G, df)

anova(reduced, full, test = "Chisq")
```



##Receiver Operating Characteristic Curve
Here we can construct the ROC curve for the full model, akin to an $R^2$ value for linear regression.  
```{r}
library(pROC) # there are other options for ROC as well
predpr <- predict(full, type = c("response"))
roccurve <- roc(ESR ~ predpr, data = plasma)
ggroc(roccurve)
auc(roccurve)
```


Here is the ROC curve for the reduced model.
```{r}
predprred <- predict(reduced, type = c("response"))
roccurve_red <- roc(ESR ~ predprred, data = plasma)
ggroc(roccurve_red)
auc(roccurve_red)
```
