---
title: "Notes 10a: Decision Trees"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", warning = FALSE, message = FALSE)

# standard packages we use
library(mosaic)
library(ggformula)
library(janitor)

# for data cleaning
library(tidyverse)

# data from ISLR book
library(ISLR2)

# package for decision trees
library(tree)

```

## Regression Trees for Baseball Data

Suppose we want to predict `Salary` based on `Years` and `Hits` for baseball players. `Salary` is recorded in thousands of dollars; `Year`s is number of years the player has played in the major leagues, and `Hits` is the number of hits the player has made in the previous year.

```{r}
#  Read in the data

data(Hitters)

hitters <- Hitters |>
  clean_names() |>    
  select(salary, years, hits) |>
  drop_na(salary) |>
  mutate(log_salary = log(salary))

head(hitters)
str(hitters)  

#fit the decision tree (regression tree)
fit_reg_tree <- tree(log_salary ~ years + hits, data = hitters)
fit_reg_tree

#Graph the regression tree
plot(fit_reg_tree)
text(fit_reg_tree, pretty = 0) #pretty=0 just tells R to print factor names directly instead of giving them code letters

#Prune the tree to 3 categories
prune_salary <- prune.tree(fit_reg_tree, best = 3)
plot(prune_salary)
text(prune_salary, pretty = 0)


```

Keep in mind that we log transformed salary, so the predictions from the tree are on the log scale. To get predictions on the original scale, we need to exponentiate the final leaves to understand the predictions. Each of these values represents thousands of dollars. 

```{r}
#Exponentiate leaves
exp(5.107)
exp(5.998)
exp(6.740)
```

## Other packages

Note there are other packages that also fit decision trees, such as `rpart`. Used with the `partykit` package, it can make some nicer diagrams, but results do not perfectly match the `tree` package. Here is an examplew with the baseball data.

```{r}
library(rpart)
library(partykit)
fit <- rpart(log_salary ~ years + hits, 
             data = hitters, 
             control = rpart.control(maxdepth = 2))
plot(as.party(fit))
```


## Linear Model Approach

Allowing for an interaction term between years and hits as well. 

```{r}

#exploratory analysis
gf_histogram(~salary, data = hitters)
#log transform
gf_histogram(~log_salary, data = hitters)

#scatterplots of salary by x variables
gf_point(log_salary ~ years, data = hitters)
gf_point(log_salary ~ hits, data = hitters)

salary_model <- lm(log_salary ~ years + hits + years*hits, data = hitters)
summary(salary_model)

#check assumptions
plot(salary_model, 1) #resids vs fitted
plot(salary_model, 2) #qq
plot(salary_model, 5) #cooks distance

#log-linear model so to get predictions on the original scale, we need to exponentiate the predictions from the model
exp(coef(salary_model))
```

