---
title: "Notes 10: Decision Trees and Random Forests"
format:
  html:
    embed-resources: true
execute: 
  warning: false
  echo: true
  error: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, comment = "", warning = FALSE, message = FALSE)

# standard packages we use
library(mosaic)
library(ggformula)
library(janitor)

# for data cleaning
library(tidyverse)

# data from ISLR book
library(ISLR2)

# package for decision trees
library(tree)

```

## Regression Trees for Baseball Data

Suppose we want to predict `Salary` based on `Years` and `Hits` for baseball players. `Salary` is recorded in thousands of dollars; `Year`s is number of years the player has played in the major leagues, and `Hits` is the number of hits the player has made in the previous year.

```{r}
#  Read in the data

data(Hitters)

hitters <- Hitters |>
  clean_names() |>    
  #select(salary, years, hits) |>
  drop_na(salary) |>
  mutate(log_salary = log(salary))

head(hitters)
str(hitters)  

#fit the decision tree (regression tree)
fit_reg_tree <- tree(log_salary ~ years + hits, data = hitters)
fit_reg_tree

#Graph the regression tree
plot(fit_reg_tree)
text(fit_reg_tree, pretty = 0) #pretty=0 just tells R to print factor names directly instead of giving them code letters

#Prune the tree to 3 terminal nodes
prune_salary <- prune.tree(fit_reg_tree, best = 3)
plot(prune_salary)
text(prune_salary, pretty = 0)


```

Keep in mind that we log transformed salary, so the predictions from the tree are on the log scale. To get predictions on the original scale, we need to exponentiate the final leaves to understand the predictions. Each of these values represents thousands of dollars. 

```{r}
#Exponentiate leaves
exp(5.107)
exp(5.998)
exp(6.740)
```

## Other packages

Note there are other packages that also fit decision trees, such as `rpart`. Used with the `partykit` package, it can make some nicer diagrams, but results do not perfectly match the `tree` package. Here is an examplew with the baseball data.

```{r}
library(rpart)
library(partykit)
fit <- rpart(log_salary ~ years + hits, 
             data = hitters, 
             control = rpart.control(maxdepth = 2))
plot(as.party(fit))
```


## Linear Model Approach

Allowing for an interaction term between years and hits as well. 

```{r}

#exploratory analysis
gf_histogram(~salary, data = hitters)
#log transform
gf_histogram(~log_salary, data = hitters)

#scatterplots of salary by x variables
gf_point(log_salary ~ years, data = hitters)
gf_point(log_salary ~ hits, data = hitters)

salary_model <- lm(log_salary ~ years + hits + years*hits, data = hitters)
summary(salary_model)

#check assumptions
plot(salary_model, 1) #resids vs fitted
plot(salary_model, 2) #qq
plot(salary_model, 5) #cooks distance

#log-linear model so to get predictions on the original scale, we need to exponentiate the predictions from the model
exp(coef(salary_model))
```

## Cross-Validation and Pruning

Normally when fitting a decision tree, it is best practice to train the model first and then use cross-validation for pruning, then fit to test data. 

First we show a large regression tree with multiple predictors for log_salary.

```{r, fig.width=6, fig.height=7}
#fit the decision tree (regression tree)
hitters_nosal <- hitters |>
  select(log_salary, years, rbi, put_outs, hits, walks, runs)

#split the data into training and testing (just 50/50)
set.seed(410)
train <- sample(1:nrow(hitters_nosal), nrow(hitters_nosal)/2)
#hitters_test will be used later for the random forest
hitters_test <- hitters_nosal[-train, "log_salary"]

#fit a regression tree on the training data. Results will not perfectly match the reading because we randomly split the data in a different way!
fit_reg_tree_large <- tree(log_salary ~ ., data = hitters_nosal, subset = train)
fit_reg_tree_large

#Graph the regression tree
plot(fit_reg_tree_large)
text(fit_reg_tree_large, pretty = 0) 
```

Cross-validation for pruning: We will still use the baseball data.
```{r}
cv_hitters <- cv.tree(fit_reg_tree_large)

#smallest deviance is desirable: size 7 means 7 terminal nodes are suggested from this analysis.
cv_hitters

plot(cv_hitters$size, cv_hitters$dev, type = "b")

#now prune to 7 terminal nodes
prune_hitters <- prune.tree(fit_reg_tree_large, best = 7)
plot(prune_hitters)
text(prune_hitters, pretty = 0)

```

## Random Forest Model

### Fit the Model

```{r}
library(randomForest)

set.seed(411)

#the "." in the model below says to consider all possible predictors in the model, rather than specifying predictors to use
#mtry should be about the square root of the number of predictors. here we have 6 predictors; we will consider 3 for each model
rf_hitters <- randomForest(log_salary ~ ., mtry = 3, data = hitters_nosal, subset = train, importance = TRUE, na.action = na.omit)
rf_hitters
```

### Predictions from Random Forest

Now we predict with the test data
```{r}

#yh_rf is the predicted log salary from each player. How could we get predicted salaries?
yhat_rf <- predict(rf_hitters, newdata = hitters_nosal[-train, ]) 

exp(yhat_rf)

#calculate SSE
mean((yhat_rf - hitters_test)^2)

```

### Variable Importance

```{r}
importance(rf_hitters)
varImpPlot(rf_hitters)

```